Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-07 ms.qmd
Re-order paragraphs

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here, we show that the relative variance ($lnCVR_{Q_{10}}$) in physiological rates for ectotherms is expected to decrease as temperatures increase, and this decrease is particularly pronouced in terrestrial ectotherms (~28%). Decreases in the relative variance of physiological rates are equal to or larger than the expected benefits from acclimation ($lnRR_{Q_{10_{acclimation}}}$) in marine and terrestrial ectotherms (*marine*: 16% acclimation vs. 15% reduction in relative variance; *terrestrial*: no acclimation vs 28% reduction in relative variance; @fig-1). Such effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide insight into the multifaceted ways in which climate warming will impact ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention. We discuss our findings in the context of the current literature and highlight key areas for future research.

@@

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial ecosystems. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model parameters to capture biologically relevant changes in variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.

>>>>

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here, we show that the relative variance ($lnCVR_{Q_{10}}$) in physiological rates for ectotherms is expected to decrease as temperatures increase, and this decrease is particularly pronouced in terrestrial ectotherms (~28%). Decreases in the relative variance of physiological rates trade-off with the expected benefits from acclimation in marine and terrestrial ectotherms (*marine*: 16% acclimation vs. 15% reduction in relative variance at high temperatures; *terrestrial*: no acclimation vs 28% reduction in relative variance at high temperatures; @fig-1). Such effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide insight into the multifaceted ways in which climate warming will impact ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention. We discuss our findings in the context of the current literature and highlight key areas for future research.

@@

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial ecosystems. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model parameters to capture biologically relevant changes in variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.
